---
title: "Assignment 2"
author: "Tenzin Tashi"
date: "April 28, 2022"
output:
  html_document:
    df_print: paged
---

#### Exercise 2 #### 

## Question a: Load the file zeta.csv of income data into R.
```{r}
setwd("/Users/tenzintashi/Downloads/CSc 460 - DS/Assignment")
# file <- file.choose()
# zeta <- read.csv(file)
zeta <- read.csv(file='Dataset/zeta.csv', header=TRUE, sep=",")
View(zeta)
```
## Question b: Change the column names of your data frame so that zcta becomes zipCode and 
## meanhouseholdincome becomes income.
``````{r}
names(zeta)[2] <- "zipCode"
names(zeta)[7] <- "income"
View(zeta)
```
## Question c: Analyze the summary of your data. 
```{r}
summary(zeta)
```
## What are the mean and median average incomes?
```{r}
# Mean average Income = 48245
# Median average Income = 44163
```
## Question d: Plot a scatter plot of the data. Although this graph is not too informative, 
## do you see any outlier values? If so, what are they?

```{r}
library(ggplot2)
ggplot(zeta,aes(x= zipCode, y=income)) +geom_point(alpha=0.2) +labs(x="Zip Code",y="Income",title="Scaterrplot Income vs Zip Code")
```
There seem to be two outlier values are 0 and 250000

## Question e: In order to omit outliers, create a subset of the data so that: $7,000 < income < $200,000, 
## What’s your new mean?
```{r}
newData <- subset(zeta, income <200000 & income >7000)
summary(newData)
```
```{r}
# New Mean average Income = 48465
```
## Question f: Create a simple box plot of your data. Be sure to add a title and label the axes.
```{r}
boxplot(col="white", data = newData, newData$income ~ newData$zipCode, main = "Average Household Income by Zip Code", xlab = "Zip Codes", ylab = "Income")
```

## Question g: In the box plot you created, notice that all of the income data is pushed towards 
## the bottom of the graph because most average incomes tend to be low. Create a new box plot 
## where the y- axis uses a log scale. Be sure to add a title and label the axes.
```{r}
#Create a new box plot where the y-axis uses a log scale
boxplot(col="white", data = newData, newData$income ~ newData$zipCode, main = "Average Household Income by Zip Code", xlab = "Zip Codes", ylab = "Income", log='y')
```
## Question h: Use the ggplot library in R, which enables you to create graphs with several different
## types of plots layered over each other. Be sure to read the documentation for ggplot
## and load the library ggplot2 (you may have to install this package into R).
```{r}
library(ggplot2)
#Make a ggplot that consists of just a scatter plot using the function geom_point() with position = “jitter” 
ggplot(newData, aes(x = as.factor(zipCode), y= income))+geom_point(position="jitter", alpha=0.2)+scale_y_log10()
```
## Question i: Make a ggplot that consists of just a scatter plot using the function geom_point()
## with position = “jitter” so that the data points are grouped by zip code. Be sure to use ggplot’s
## function for taking the log10 of the y-axis data. (Hint: for geom_point, have alpha=0.2).
```{r}
library(ggplot2)
ggplot(newData, aes(x = as.factor(zipCode), y = income))+geom_point(position = 'jitter',alpha=0.2)+geom_boxplot(alpha=0.1,outlier.size =-Inf) + scale_y_log10()+labs(color="Region",x="Zip Code",y="Income",title="Average Income by Zip Code") + theme(plot.title = element_text(size =11, face="plain",hjust = 0.5))
```
## Question j: Create a new ggplot by adding a box plot layer to your previous graph.
## To do this, add the ggplot function geom_boxplot(). Also, add color to the scatter plot so
## that data points between different zip codes are different colors. Be sure to label the axes and
## add a title to the graph. (Hint: for geom_boxplot, have alpha=0.1 and outlier.size=0).
```{r}
library(ggplot2)
ggplot(newData,aes(x=as.factor(zipCode),y=income))+geom_point(aes(colour=factor(zipCode)),position = 'jitter',alpha=0.2)+ geom_boxplot(alpha=0.1,outlier.size =0) + scale_y_log10()+ ylab("Income") + xlab("Zip Code") + ggtitle ("Average Income by Zip Code") + labs(color="Region") + theme(plot.title = element_text(size =11, face="plain",hjust = 0.5))
```
## Question k: What can you conclude from this data analysis/visualization?
```{r}
# - It is important to visualize your data in different ways. 

# - Visualization enables you to better understand what your data is telling you. 

# - Visualization enables you to better communicate your results to stakeholders. 

# - Zip codes starting in 0 (New England) and 9 (West Coast) have higher average household incomes. 
```


#### Exercise 2.2 #### 

## Question a: Cluster the data and plot all 52 data points, along with the centroids.
## Mark all data points and centroids belonging to a given cluster with their own color. Here, let k=10.
```{r}
load('income_elec_state.Rdata')
k = kmeans(income_elec_state, 10)
plot(income_elec_state, col = k$cluster)
points(k$centers, col=1:10, pch=8)
```
## Question b: Repeat step (a) several times. What can change each time you cluster the data?
## Why? How do you prevent these changes from occurring?
```{r}
k = kmeans(income_elec_state, 10)
plot(income_elec_state, col = k$cluster)
points(k$centers, col=1:10, pch=8)
```

```{r}
k = kmeans(income_elec_state, 10)
plot(income_elec_state, col = k$cluster)
points(k$centers, col=1:10, pch=8)
```

```{r}
k = kmeans(income_elec_state, 10)
plot(income_elec_state, col = k$cluster)
points(k$centers, col=1:10, pch=8)
```

```{r}
# Sizes, centers’ position, sum of squares of clusters can change after each time repeat above step.
# Because by default nstart = 1: having only one random starting set can result in different
# clusterings over multiple runs.

# To prevent these changes from occurring, we can:
  # - Increase “nstart” to improve the likelihood of obtaining the globally optimal clustering.
  # - Increasing the “iter.max” parameter reduces the likelihood that the kmeans algorithm terminates     
  #   prematurely.
```

```{r}
k = kmeans(income_elec_state, 10, nstart=100, iter.max = 50)
plot(income_elec_state, col = k$cluster)
points(k$centers, col=1:10, pch=8)
```
## Question c: Once you’ve accounted for the issues in the previous step, determine a reasonable 
## value of k. Why would you suggest this value of k?
```{r}
wss =numeric(10)
for (i in 1:10) wss[i] = sum(kmeans(income_elec_state, centers=i, nstart = 100, iter.max = 50)$tot.withinss)
plot(1:10, wss, type="b", xlab="Number of Clusters", ylab="Total within-clusters sum of squares")
```
We can see here "elbow" = 4


With k=4

```{r}
k = kmeans(income_elec_state, 4, nstart=100, iter.max = 50)
plot(income_elec_state, col = k$cluster)
points(k$centers, col=1:4, pch=8)
```
Repeat the modeling with k=3

```{r}
k = kmeans(income_elec_state, 3, nstart=100, iter.max = 50)
plot(income_elec_state, col = k$cluster)
points(k$centers, col=1:4, pch=8)
```
Repeat the modeling with k=5

```{r}
k = kmeans(income_elec_state, 5, nstart=100, iter.max = 50)
plot(income_elec_state, col = k$cluster)
points(k$centers, col=1:4, pch=8)
```

Chosen k=4. Because we see that Puerto Rico is an outlier, and should perhaps belong to its own cluster. It is the smallest k such that Puerto Rico
belongs to its own cluster, so this k would be a good value to suggest.


## Question d: Convert the mean household income and mean electricity usage to a log10 scale and
## cluster this transformed dataset. How has the clustering changed? Why?
```{r}
new = log10(income_elec_state)
k = kmeans(new, 10, nstart=100, iter.max = 50)
plot(new, col = k$cluster)
points(k$centers, col=1:10, pch=8)
```
K-means clustering is not scale-invariant, so any adjustments made to the units of the data may impact the clustering.

## Question e: Reevaluate your choice of k. Would you now choose k differently? Why or why not?
```{r}
wss =numeric(10)
for (i in 1:10) wss[i] = sum(kmeans(new, centers=i, nstart = 100, iter.max = 50)$tot.withinss)
plot(1:10, wss, type="b", xlab="Number of Clusters", ylab="Total within-clusters sum of squares")
```
We see more clear elbow in the different position: k=5

## Question f: Have you observed an outlier in the data? Remove the outlier and, once again, reevaluate your
## choice of k
```{r}
new <- subset(new, rownames(new) != "PR")

wss =numeric(10)
for (i in 1:10) wss[i] = sum(kmeans(new, centers=i, nstart = 100, iter.max = 50)$tot.withinss)
plot(1:10, wss, type="b", xlab="Number of Clusters", ylab="Total within-clusters sum of squares")
```
After removing the outliers, it is clear that elbow on the plot change its position to smaller value. k=4

## Question g: Color a map of the U.S. according to the clustering you obtained. To simplify this task, use 
## the “maps” package and color only the 48 contiguous states and Washington D.C.
```{r}
library(maps)

km <- kmeans(new,4,nstart = 100, iter.max = 50)
#Prepare vector with state order
map_order <- c('AL', 'AZ', 'AR', 'CA', 'CO', 'CT', 'DE', 'DC', 'FL',
'GA', 'ID', 'IL', 'IN', 'IA', 'KS', 'KY', 'LA', 'ME',
'MD', 'MA', 'MA', 'MA', 'MI', 'MI', 'MN', 'MS', 'MO',
'MT', 'NE', 'NV', 'NH', 'NJ', 'NM', 'NY', 'NY', 'NY',
'NY', 'NC', 'NC', 'NC', 'ND', 'OH', 'OK', 'OR', 'PA',
'RI', 'SC', 'SD', 'TN', 'TX', 'UT', 'VT', 'VA', 'VA',
'VA', 'WA', 'WA', 'WA', 'WA', 'WA', 'WV', 'WI', 'WY')
#Prepare color vector
map_color <- km$cluster[map_order]
map('state', col = map_color,fill=TRUE)
```
